Validation of Visual Statistical Inference, with Application to Linear Models

### 2012-03-04 paper_jasa.tex
plot signal strength added
trying to clean up and tighten the introduction

The $(m-1)$ null plots can be considered to be samples drawn from the sampling distribution of the the test statistic assuming that the null hypothesis is true. If $H_1$ is true, the most different plot is expected to be the test statistic, i.e, the plot of the data. {\em If the test statistic cannot be identified} in the lineup the conclusion is to {\em not reject the null hypothesis.}  A careful visual insfection follows, and the observer is asked to point out the plot most different from the lineup.

@@

The lineup plot can be evaluated by one or more individuals. When a single individual identifies the observed graph in the lineup plot we report a $p$-value of at most $1/m$, otherwise the $p$-value is at least $1-\frac 1m$.

If $N$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (N,\frac{1}{m})$ and we report a  $p$-value of at most $Pr(U \ge u)= \sum_{k \ge u}^N {{N \choose k} (\frac{1}{m})^k(1-\frac 1m)^{(N-k)}}$ where $u$ is the observed number of successful evaluations. %Notice that when $N=1$, this $p$-value is $\frac1m$.

@@

For two different visual test statistics of the same observed data, the one  is better, in which a specific pattern is more easily distinguishable visually. This should be reflected in the power of the test. We can assess power therefore both empirically based on experimental data and through theoretical considerations. Next, we will develop power theoretically and then relate it to the empirical results.

\begin{dfn} \label{dfn:power}
For a lineup of $m$ plots the power of $\theta$ is defined as 
    \begin{equation*}
      \text{Power}(\theta)= 
        \begin{cases} 
              \text{Type-I error}=\frac1m & \text{if $\theta \in \Theta_0$,} \\
              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
        \end{cases}
    \end{equation*}
\end{dfn}

Power has a lower limit of $\frac1m$ since the probability that a person will randomly pick the test statistic under $H_0$ is $\frac1m$. In general,  $Pr(\text{Reject } H_0)$ is determined by the type of test being conducted. Theoretical power for the regression parameters is derived in the next section.

\section{Inference for a Regression Model} \label{sec:regression}

@@

In a regression setting, there is a multitude of established graphics that help with the evaluation and diagnostics of a model (e.g. Cook \& ... ). Table \ref{tbl:stat_multiple} shows an overview of some commonly used charts and their corresponding hypothesis tests. Note that some charts can be associated with several different testing situations. In our paper, we pick the example fo situation 2 and 3 as candidates.

A lineup including this test statistic is shown in Figure \ref{fig:test_category}. The 19 null plots are generated  by simulating residuals from $N(0,\hat{\sigma^2})$. The test statistic, the plot containing the observed data, is randomly placed among these null plots. If the test statistic is identifiable the null hypothesis is rejected with a $p$-value of at most 0.05.

@@

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
Now consider estimating the power of the visual test. Suppose $F_{|t|}(.)$ be the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in classical test be $t_{obs}$ with $p$-value $p_B$. By definition we have

@@

\subsection{Expected number of choices}
Since $X$ has a binomial distribution, we can have a look at the number of plots among the null plots that we would expect to be picked over the data plot:
\[
E[X] = (m-1)(1-E[p_B]).
\]
With an increase of the lineup size $m$ the number of null plots with a potentially stronger signal than the data plot grows linearly.
This should allow us to infer some of the signal strength $p_B$ as
\[
E[p_B] = 1 - E[X]/(m-1),
\]
i.e. by averaging over the number of plots with  a stronger signal than the data plot we can evaluate signal strength of the plot even in the case where the data plot is not being picked by the observer.

@@

>>>>

The $(m-1)$ null plots can be considered to be samples drawn from the sampling distribution of the test statistic assuming that the null hypothesis is true.
If $H_1$ is true, \blue{we expect this to be reflected as a feature in the test statistic, i.e. the plot of the data, that makes it visually distinguishable from the null plots.}
{\em If the test statistic cannot be identified} in the lineup the conclusion is to {\em not reject the null hypothesis.}  A careful visual inspection follows, and the observer is asked to point out the plot most different from the lineup.

@@

%The lineup plot can be evaluated by one or more individuals. When a single individual identifies the observed graph in the lineup plot we report a $p$-value of at most $1/m$, otherwise the $p$-value is at least $1-\frac 1m$.

%If $N$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (N,\frac{1}{m})$ and we report a  $p$-value of at most $Pr(U \ge u)= \sum_{k \ge u}^N {{N \choose k} (\frac{1}{m})^k(1-\frac 1m)^{(N-k)}}$ where $u$ is the observed number of successful evaluations. %Notice that when $N=1$, this $p$-value is $\frac1m$.

@@

For two different visual test statistics of the same observed data, the one  is better, in which a specific pattern is more easily distinguishable to the observer. This should be reflected in the power of the test. We can assess power therefore both empirically based on experimental data and through theoretical considerations. 
\blue{

\begin{dfn} \label{dfn:test}
A \textit{visual test} $T_\theta$ is defined as the grammar \citep{wilkinson:1999} of the plot used in the lineup. The grammar consists of type and specification of aesthetics necessary for complete reproducibility.
\end{dfn}
}
Next, we will develop power theoretically and then relate it to the empirical results.

\begin{dfn} \label{dfn:power}
\red{HH: I don't think power should be defined in terms of $\theta$ Usually power is considered a property of the test.  In our case that's the specifics of the plot used for the lineup. We might need an additional definition for that - maybe we can phrase that within the framework of the grammar of graphics? - that's what I've tried above}

For a lineup of $m$ plots the power of \blue{a visual test}
%$\theta$
 is defined as 
    \begin{equation*}
      \text{Power}(T_\theta)= 
        \begin{cases} 
              \text{Type-I error}=\frac1m & \text{if $\theta \in \Theta_0$,} \\
              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
        \end{cases}
    \end{equation*}
\end{dfn}
\red{this is a non-standard definition of power - usually power is just the prob to reject under $H_1$. could we back the non-standard definition up with some literature?}

\red{HH: let's try to keep the power discussion as general as possible, and only use regression as an example.}

For any single observer, the power of a visual test has a lower limit of $\frac1m$ since the probability that an observer will randomly pick the test statistic under $H_0$ is $\frac1m$.

% HH: I think this is covered by making power a property of the test
% In general,  $Pr(\text{Reject } H_0)$ is determined by the type of test being conducted.

\red{HH: We need to define p values corresponding to the data shown in plots. }
\red{second step: the p-values under $H_0$ are uniform.}

Theoretical power for the regression parameters is derived in the next section.
\section{Inference for a Regression Model} \label{sec:regression}

@@

In a regression setting, there is a multitude of established graphics that help with the evaluation and diagnostics of a model \citep{cook:99}. Table \ref{tbl:stat_multiple} shows an overview of some commonly used charts and their corresponding hypothesis tests. Note that some charts can be associated with several different testing situations. In our paper, we pick the examples for situation 2 and 3 as candidates.

A lineup including this test statistic is shown in Figure \ref{fig:test_category}. The 19 null plots are generated  by simulating residuals from $N(0,\hat{\sigma^2})$. The test statistic, the plot containing the observed data, is randomly placed among these null plots. If the test statistic is identifiable the null hypothesis is rejected. % with a $p$-value of at most 0.05.

@@

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
Now consider estimating the power of the visual test. 
Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in classical test be $t_{obs}$ with $p$-value $p_B$. By definition we have

@@

\subsection{Expected number of choices}
Since $X$ has a binomial distribution, we can have a look at the number of plots among the null plots that we would expect to be picked over the data plot:
\[
E[X] = (m-1)(1-E[p_B]).
\]
With an increase of the lineup size $m$ the number of null plots with a potentially stronger signal than the data plot grows linearly.
This should allow us to infer some of the signal strength $p_B$ as
\begin{equation} \label{plot_signal}
E[p_B] = 1 - E[X]/(m-1),
\end{equation}
i.e. by averaging over the number of plots with  a stronger signal than the data plot we can evaluate signal strength of the plot even in the case where the data plot is not being picked by the observer.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{p-value vs plot signal strength. The solid line represents 45 degree line which represents the equity shown in equation \eqref{plot_signal}}
       \label{fig:pval_plot_signal}
\end{figure*}
